# Working with data structures

## Vectors

In the first exercise, you started to explore **data structures**, which store information about data types. You played around with **vectors**, which is a ordered collection of a data type. Each *element* of a vector contains a data type, and there is no limit on how big a vector can be, as long the memory use of it is within the computer's memory (RAM).

We can now store a vast amount of information in a vector, and assign it to a single variable. We can now use operations and functions on a vector, modifying many elements within the vector at once! This fits with the feature of "encapsulate complex data via data structures to allow efficient manipulation of data" described in the first lesson!

We often create vectors using the combine function, `c()` :

```{r}
staff = c("chris", "shasta", "jeff")
chrNum = c(2, 3, 1)
```

If we try to create a vector with mixed data types, R will try to make them be the same data type, or give an error:

```{r}
staff = c("chris", "shasta", 123)
staff
```

Our numeric got converted to character so that the entire vector is all characters.

### Using operations on vectors

Recall from the first class:

-   Expressions are be built out of **operations** or **functions**.

-   Operations and functions combine **data types** to return another data type.

Now that we are working with data structures, the same principle applies:

-   Operations and functions combine **data structures** to return another data structure (or data type!).

What happens if we use some familiar operations we used for numerics on a numerical vector? If we multiply a numerical vector by a numeric, what do we get?

```{r}
chrNum = chrNum * 3
chrNum 
```

All of `chrNum`'s elements tripled! Our multiplication operation, when used on a *numeric vector with a numeric*, has a *new* meaning: it multiplied all the elements by 3. Multiplication is an operation that can be used for multiple data types or data structures: we call this property **operator overloading**. Here's another example: *numeric vector multiplied by another numeric vector*:

```{r}
chrNum * c(2, 2, 0)
```

but there are also limits: a numeric vector added to a character vector creates an error:

```{r}
#chrNum + staff
```

When we work with operations and functions, we must be mindful what inputs the operation or function takes in, and what outputs it gives, no matter how "intuitive" the operation or function name is.

### Subsetting vectors explicitly

In the exercise this past week, you looked at a new operation to subset elements of a vector using brackets.

Inside the bracket is either a single numeric value or an a **numerical indexing vector** containing numerical values. They dictate which elements of the vector to return.

```{r}
staff[2]
staff[c(1, 2)]
small_staff = staff[c(1, 2)]
```

In the last line, we created a new vector `small_staff` that is a subset of the staff given the indexing vector `c(1, 2)`. We have three vectors referenced in one line of code. This is tricky and we need to always refer to our rules step-by-step: evaluate the expression right of the `=`, which contains a vector bracket. Follow the rule of the vector bracket. Then store the returning value to the variable left of `=`.

Alternatively, instead of using numerical indexing vectors, we can use a **logical indexing vector**. The logical indexing vector must be the *same length* as the vector to be subsetted, with `TRUE` indicating an element to keep, and `FALSE` indicating an element to drop. The following block of code gives the same value as before:

```{r}
staff[c(TRUE, FALSE, FALSE)]
staff[c(TRUE, TRUE, FALSE)]
small_staff = staff[c(TRUE, TRUE, FALSE)]
```

### Subsetting vectors implicitly

Here are two applications of subsetting on vectors that need distinction to write the correct code:

1.  **Explicit subsetting**: Suppose someone approaches you a 100-length vector of people's ages, and say that they want to subset to the first 10 elements.

2.  **Implicit subsetting**: Suppose someone approaches you a 100-length vector of people's ages, and say that they want to subset to elements \< 18 age.

We already know how to explicitly subset:

```{r}
set.seed(123) #don't worry about this function
age = round(runif(100, 1, 100)) #don't worry about these functions
first_ten_age = age[1:10]
```

For implicit subsetting, we don't know which elements to select off the top of our head! If we know which elements have less than 18, then we can give the elements for an explicit subset. Therefore, we need to create a logical indexing vector using a **comparison operator**:

```{r}
indexing_vector = age < 18
indexing_vector
```

The comparison operator `<` compared the numeric value of `age` to see which elements of age is less than 18, and then returned a logical vector that has `TRUE` if age is less than 18 at that element and `FALSE` otherwise.

Then,

```{r}
age_young = age[indexing_vector]
age_young
```

We could have done this all in one line without storing the indexing vector as a variable in the environment:

```{r}
age_young = age[age < 18]
```

We have the following comparison operators in R:

`<` less than

`<=` less or equal than

`==` equal to

`!=` not equal to

`>` greater than

`>=` greater than or equal to

You can also put these comparison operators together to form more complex statements, which you will explore in this week's exercise.

Another example:

```{r}
age_90 = age[age == 90]
age_90
age_not_90 = age[age != 90]
age_not_90
```

For most of our subsetting tasks on vectors (and dataframes below), we will be encouraging implicit subsetting. The power of implicit subsetting is that you don't need to know what your vector contains to do something with it! This technique is related to *abstraction* in programming mentioned in the first lesson: by using expressions to find the specific value you are interested instead of *hard-coding* the value explicitly, it generalizes your code to handle a wider variety of situations.

## Dataframes

Before we dive into dataframes, check that the `tidyverse` package is properly installed by loading it in your R Console:

```{r, message=F}
library(tidyverse)
```

Here is the data structure you have been waiting for: the **dataframe**. A dataframe is a spreadsheet such that each column must have the same data type. Think of a bunch of vectors organized as columns, and you get a dataframe.

For the most part, we load in dataframes from a file path (although they are sometimes created by combining several vectors of the same length, but we won't be covering that here):

```{r}
load(url("https://github.com/fhdsl/S1_Intro_to_R/raw/main/classroom_data/CCLE.RData"))
```

### Using functions and operations on dataframes

We can run some useful functions on dataframes to get some useful properties, similar to how we used `length()` for vectors:

```{r}
nrow(metadata)
ncol(metadata)
dim(metadata)
colnames(metadata)
```

The last function, `colnames()` returns a character vector of the column names of the dataframe. This is an important property of dataframes that we will make use of to subset on it.

We introduce an operation for dataframes: the `dataframe$column_name` operation selects for a column by its column name and returns the column as a vector. For instance:

```{r}
metadata$OncotreeLineage[1:5]
metadata$Age[1:5]
```

We treat the resulting value as a vector, so we can perform implicit subsetting:

```{r}
metadata$OncotreeLineage[metadata$OncotreeLineage == "Myeloid"]
```

The bracket operation `[ ]` on a dataframe can also be used for subsetting. `dataframe[row_idx, col_idx]` subsets the dataframe by a row indexing vector `row_idx`, and a column indexing vector `col_idx`. 

```{r}
metadata[1:5, c(1, 3)]
```

We can refer to the column names directly:

```{r}
metadata[1:5, c("ModelID", "CellLineName")]
```


We can leave the column index or row index empty to just subset columns or rows.

```{r}
metadata[1:5, ]
```


```{r}
head(metadata[, c("ModelID", "CellLineName")])
```

The bracket operation on a dataframe can be difficult to interpret because multiple expression for the row and column indicies is a lot of information for one line of code. You will see easier-to-read functions for dataframe subsetting in the next lesson.

Lastly, try running `View(metadata)` in RStudio Console...whew, a nice way to examine your dataframe like a spreadsheet program!

# Working with Data Structures Exercises

If you need to clear your environment and start from the beginning, this code chunk will do that:

```{r}
rm(list = ls())
```

## Recap from class: subsetting explicitly and implicitly

In class, we looked at a numerical vector similar to this:

```{r}
age = c(89, 70, 64, 99, 66, 71, 55, 60, 30, 16)
```

We could subset `age` **explicitly** two ways. Suppose we want to subset the 1st and 5th, and 9th elements. One can do it with numerical indexing vectors:

```{r}
age[c(1, 5, 9)]
```

or by **logical indexing vectors**:

```{r}
age[c(TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE)]
```

and you can do it in one step as we have done so, or two steps by storing the indexing vector as a variable. *Either ways is fine.*

```{r}
num_idx = c(1, 5, 9)
age[num_idx]
```

```{r}
logical_idx = c(TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE)
age[logical_idx]
```

Now, let's subset this vector **implicitly**, in 3 steps:

1.  Come up with a criteria for subsetting: "I want to subset to values greater than 50".
2.  We can use a **comparison operator** to create a **logical indexing vector** that fits this criteria.

```{r}
age > 50
```

3.  Use this logical indexing vector to subset.

```{r}
age[age > 50]
```

And you are done.

Alternatively, you could have done this in two steps:

```{r}
another_idx = age > 50
age[another_idx]
```

On to practicing yourself!

## Part 1: Comparison operators on data types

In class, we looked at comparison operators on vectors to create logical indexing vectors. We are going to first take a step back and see how comparison operators work on simpler data types. Here are the comparison operators common in R:

`<` less than

`<=` less or equal than

`==` equal to

`!=` not equal to

`>` greater than

`>=` greater than or equal to

Consider the following variables:

```{r}
chr1 = 2
chr2 = 3
chr3 = 1
staff1 = "chris"
staff2 = "shasta"
```

We saw how we used comparison operators on vectors, but they can be used for any singular data type:

Using comparison operators, write 3 expressions so that the returned value is `TRUE`. I'll give one example:

```{r}
staff1 != "jeff"


```

Using comparison operators, write 3 expressions so that the returned value is `FALSE`:

```{r}
chr1 > 2


```

We can combine comparison operators together to form more complex statements:

The "and" `&` operator returns `TRUE` if both inputs have values of `TRUE`, and `FALSE` otherwise:

```{r}
a = TRUE
b = TRUE
c = FALSE
d = FALSE
a & b
c & d
a & c
b & d
```

Therefore, we can create compound comparison operations like this:

```{r}
chr1 > chr3 & chr1 < chr2
```

Similarly, the "or" `|` operator returns `TRUE` if at least one of the inputs are `TRUE`. Otherwise it returns `FALSE`.

```{r}
a | b
c | d
a | c
b | d
```

Create 2 compound comparison operations using `|` so that one returns `TRUE` and the other returns `FALSE`. I'll give you one example:

```{r}
chr1 > chr3 | chr1 < chr2

```

You can extend these compound statements to contain multiple `&` and `|` operations mixed together, and parentheses should be used to clarify your order of operation.

**Now, let's use comparison operators on vectors.**

We create the `age` vector from our lesson.

```{r}
set.seed(123) #don't worry about this function
age = round(runif(20, 1, 100)) #don't worry about these functions
```

Create a compound comparison operation that looks returns a logical vector so that the elements are `TRUE` if `age` is between 18 *and* 65.

```{r}

```

## Part 2: Subsetting vectors

Let's practice explicit subsetting. Can you subset for the first 10 elements of `age` explicitly?

```{r}

```

We will now look at implicit subsetting using with a criteria of interest: "We want to subset `age` to have values less than 21 *or* greater than 40".

Then, the next step is to create the logical indexing vector via compound comparison operations. You should use the or `|` operation:

```{r}

```

The final step is to use this logical indexing vector to subset:

```{r}

```

Let's do another practice: Can you subset for all `age` greater than 65?

```{r}

```

## Part 3: Dataframes

```{r, message=F, warning=F}
library(tidyverse)
load(url("https://github.com/fhdsl/Intro_to_R/raw/main/classroom_data/CCLE.RData"))
```

Use the `colnames`, `dim`, `ncol`, and `nrow` functions to explore the basic properties of the dataframe `metadata`:

```{r}

```

Using the `$` operator, subset `OncotreeLineage` column from the `metadata` dataframe and store the result as a character vector called `lineage`:

```{r}

```

Then, subset for the first 10 elements of `lineage` explicitly:

```{r}

```

Create a logical indexing vector that gives `TRUE` if an element of `lineage` is "Myeloid", and `FALSE` otherwise. You will need to use the `==` comparison operation. Store it as the variable `myeloid_idx`.

```{r}

```

We can now use this logical indexing vector to subset on the *rows* of `metadata`. Take a look (you need to remove the # comments the code):

```{r}
#myeloid_metadata = metadata[myeloid_idx ,]
#View(myeloid_metadata)
```

## You are done!

Change the logical value of this variable to `TRUE`. That's all you need to know about **logicals** for now!

```{r}
finished_with_exercise1 = FALSE
```

## Feedback!

How many hours did you spend on this exercise?

```{r}
time_spent = 0 
```

If you worked with other peers, write their names down in the following character vector: each element is one person's name.

```{r}
peers = c("Myself")
```

