---
title: "W2: Working with data structures"
format: revealjs
  #smaller: true
  #scrollable: true
execute:
  echo: true
output-location: fragment
---

## A programming language has following elements

. . .

-   Grammar structure to construct expressions

. . .

-   Combining expressions to create more complex expressions

. . .

-   Encapsulate complex expressions via functions to create modular and reusable tasks

. . .

-   Encapsulate complex data via data structures to allow efficient manipulation of data 

## Data structures

**Data structures** stores information about data types.

. . .

**Vector** is a ordered collection of a data type. Each *element* of a vector contains a data type, and all elements of a vector must be the same type, such as numeric, character, or logical.

::: notes
There is no limit on how big a vector can be, as long the memory use of it is within the computer's memory (RAM).

We can now store a vast amount of information in a vector, and assign it to a single variable. We can now use operations and functions on a vector, modifying many elements within the vector at once!
:::

## Vector example

. . .

```{r}
staff = c("chris", "shasta", "jeff")
chrNum = c(2, 3, 1)
```

. . .

If we try to create a vector with mixed data types, R will try to make them be the same data type:

```{r}
staff = c("chris", "shasta", 123)
staff
```

. . .

😱

## Using operations on vectors

. . .

Operations and functions combine **data structures** to return another data structure (or data type!).

. . .

```{r}
chrNum = chrNum * 3
chrNum 
```

. . .

Multiplication has a new meaning when used on a numeric vector and numeric data type!

. . .

😮

## Operator overloading

Multiplication is an operation that can be used for multiple data types or data structures: we call this property **operator overloading**.

. . .

How about numeric vector multiplied by numeric vector?

```{r}
chrNum * c(2, 2, 0)
```

. . .

a numeric vector added to a character vector:

```{r}
#chrNum + staff
```

## Subsetting vectors

We subset vectors using the bracket `[ ]` operation.

. . .

Inside the bracket is either a single numeric value or an a **numerical indexing vector** containing numerical values. They dictate which elements of the vector to return.

. . .

```{r}
staff[2]
```

. . .

```{r}
staff[c(1, 2)]
```

. . .

```{r}
chrNum = rep(2, 22)
chrNum[1:10]
```

## Subsetting vectors

Alternatively, instead of using numerical indexing vectors, we can use a **logical indexing vector**.

. . .

The logical indexing vector must be the *same length* as the vector to be subsetted, with `TRUE` indicating an element to keep, and `FALSE` indicating an element to drop.

. . .

```{r}
staff[c(TRUE, FALSE, FALSE)]
```

. . .

```{r}
staff[c(TRUE, TRUE, FALSE)]
```

. . .

```{r}
small_staff = staff[c(TRUE, TRUE, FALSE)]
small_staff
```

## Explicit vs. implicit subsetting

. . .

-   **Explicit subsetting**: Given a 100-length vector of people's ages, and subset to the first 10 elements.

. . .

```{r}
set.seed(123)
age = round(runif(100, 1, 100))
first_ten_age = age[1:10]
first_ten_age
```

. . .


-   **Implicit subsetting**: Given a 100-length vector of people's ages, and subset to elements \< 18 age.


## Implicit subsetting

We don't know which elements to subset off the top of our head! If we know which elements have less than 18, then we can give the elements for an explicit subset.

. . .

Use a **comparison operator** to create a logical indexing vector:

. . .

```{r}
logical_indexing_vector = age < 18
logical_indexing_vector
```

## Implicit subsetting

```{r}
logical_indexing_vector = age < 18
age_young = age[logical_indexing_vector]
age_young
```

## Other comparison operators:

`<` less than

`<=` less or equal than

`>` greater than

`>=` greater than or equal to

`==` equal to

`!=` not equal to

. . .

```{r}
staff != "chris"
```

. . .

```{r}
staff[staff != "chris"]
```


## Functions on vectors

How many elements are in this vector?

```{r}
length(staff)
```

## Dataframes

. . .

A **dataframe** is a spreadsheet such that each column must have the same data type; a bunch of vectors organized as columns.

. . .

```{r}
load(url("https://github.com/fhdsl/Intro_to_R/raw/main/classroom_data/CCLE.RData"))
nrow(metadata)
```

. . .

```{r}
ncol(metadata)
```

. . .

```{r}
dim(metadata)
```

. . .

`head(metadata)`

`View(metadata)`

## Column names of dataframes

The **column name** is a character vector that corresponds to the columns of the dataframe.

. . .

```{r}
colnames(metadata)
```

. . .

## Subsetting a column from a dataframe

. . .

The `dataframe$column_name` operation selects for a column by its column name and returns the column as a vector:

. . .

```{r}
metadata$OncotreeLineage[1:5]
```

. . .

```{r}
metadata$Age[1:5]
```

. . .

Challenge!

```{r}
metadata$OncotreeLineage[metadata$OncotreeLineage == "Myeloid"]
```

## Subsetting columns and rows from a dataframe

`dataframe[row_idx, col_idx]` subsets the dataframe by a row indexing vector `row_idx`, and a column indexing vector `col_idx`. 

. . .

```{r}
metadata[1:5, c(1, 3)]
```

. . .

We can refer to the column names directly:

```{r}
metadata[1:5, c("ModelID", "CellLineName")]
```


## Subsetting columns and rows from a dataframe


We can leave the column index or row index empty to just subset columns or rows.

. . .

```{r}
metadata[1:5, ]
```

. . .

```{r}
metadata[, c("ModelID", "CellLineName")]
```



## Appendix: Subsetting vs. Modifying part of a vector

. . .

```{r}
age = c(10, 35, 24, 70, 84)
age_young = age[age < 18]
age_young
```

. . .

```{r}
medicaid_eligible = age

medicaid_eligible[medicaid_eligible < 65] = 0
medicaid_eligible[medicaid_eligible >= 65] = 1
medicaid_eligible
```
